---
title: "Fastmix New"
author: "Meherab"
date: "2026-02-10"
output: pdf_document
---
# Intro


We want to run a version of Fastmix pipeline that can accomodate general longitudinal datasets.
At first we will use a simple DGP with 1 covariate and a random effect of the intercept and fit a model accordingly.

Here we install libraries

```{r installing_libraries, message=FALSE}

library(tidyverse)
library(lmerTest)
library(fastmap)
library(furrr)   # for future_map2
library(future)

# Uncomment and run to install fastmix (not needed at the moment)
#library(devtools)
#devtools::install_github("terrysun0302/FastMix", build_vignettes = FALSE, force = TRUE)
library(FastMix)
library(Matrix)
library(mclust)
library(purrr)
library(nortest)
library(pbapply)
library(future.apply)
library(progressr)

```


# Sub Functions

These are subfunctions that we use carry out specific operations

```{r sub_func}

###############################################
# Stable matrix inverse using SVD
# ---------------------------------------------
# This function computes a numerically stable
# inverse by truncating small singular values.
# Used to avoid instability when matrices are
# ill-conditioned (very common in LMMs).
###############################################

rsolve = function (X, min.cond.num = 1e-06) 
{
    # Singular Value Decomposition
    # X = U D V^T
    o <- svd(X)
    
    # Threshold for smallest allowable eigenvalue
    lambda <- min.cond.num * max(o$d)
    
    # Truncate small singular values
    dd <- pmax(o$d, lambda)
    
    # Invert singular values
    if (length(dd) == 1) {
        dd.inv <- 1/dd
    } else {
        dd.inv <- diag(1/dd)
    }
    
    # Return stable inverse: V D^{-1} U^T
    return(o$v %*% dd.inv %*% t(o$u))
}



#########################################################
# subject_fun()
# -------------------------------------------------------
# Computes subject-specific quantities needed for
# generalized least squares (GLS) estimation:
#
# - Marginal covariance of Y_i by calculating V_i
# - Inverse covariance using Woodbury identity
# - Components for global WLS estimation of beta
#########################################################

subject_fun <- function(i, Y, Des, DZ, Sigma_gamma, sigma2, idx_list,
                        min.cond.num = 1e-6) {

  # Indices for subject i
  idx <- idx_list[[i]]
  
  # Subject-specific response vector
  Yi <- Y[idx]
  
  # Fixed-effects design matrix for subject i
  Xi <- as.matrix(Des[idx, -1])
  
  # Random-effects design matrix for subject i
  Zi <- as.matrix(DZ[idx, -1])
  
  # Number of observations (timepoints) for subject i
  ni <- nrow(Zi)

  # Marginal covariance of Y_i: WE DO NOT USE THIS SO IT IS COMMENTED OUT
  # Var(Y_i) = Z_i Σ_γ Z_i^T + σ² I
  #Sigma_Yi <- Zi %*% Sigma_gamma %*% t(Zi) + sigma2 * diag(ni)

  #################################################
  # Woodbury identity to compute Σ_Yi^{-1}
  #
  # (σ²I + Z Σ Z^T)^{-1}
  # = (1/σ²)I − (1/σ⁴) Z (Σ^{-1} + (1/σ²)Z^TZ)^{-1} Z^T
  #################################################

  # Inner matrix in Woodbury identity
  inner <- rsolve(Sigma_gamma) + (1 / sigma2) * (t(Zi) %*% Zi)
  
  # Invert inner matrix
  W <- rsolve(inner, min.cond.num = min.cond.num)

  # Inverse of marginal covariance of Y_i
  Vi_inv <- (1 / sigma2) * diag(ni) -
            (1 / sigma2^2) * Zi %*% W %*% t(Zi)

  # Components for global GLS / WLS estimation of beta
  # p1 = X_i^T V_i^{-1} X_i
  # p2 = X_i^T V_i^{-1} Y_i
  p1 <- t(Xi) %*% Vi_inv %*% Xi
  p2 <- t(Xi) %*% Vi_inv %*% Yi

  # Return all subject-level quantities
  list(
    #Sigma_Yi = Sigma_Yi,
    Vi_inv = Vi_inv,   
    p1 = p1,
    p2 = p2,
    Xi = Xi,
    Zi = Zi,
    Yi = Yi,
    idx = idx
  )
}



#########################################################
# gamma_fun()
# -------------------------------------------------------
# Computes the conditional (posterior) distribution of
# subject-specific random effects:
#
# - E[γ_i | Y_i]  (BLUP / EBLUP)
# - Var(γ_i | Y_i)
# - Second moment E[γ_i γ_i^T | Y_i]
#
# This corresponds to the E-step in an EM algorithm.
#########################################################

gamma_fun <- function(i, subject_results, Y, Des, beta, Sigma_gamma) {

  # Extract subject-specific quantities
  Zi <- subject_results[[i]]$Zi
  Vi_inv <- subject_results[[i]]$Vi_inv
  idx <- subject_results[[i]]$idx

  # Residual after removing fixed effects
  # r_i = Y_i − X_i β
  resid_i <- Y[idx,1] - as.matrix(Des[idx, -1]) %*% beta

  #################################################
  # Posterior mean of random effects:
  #
  # E[γ_i | Y_i] =
  # Σ_γ Z_i^T V_i^{-1} (Y_i − X_i β)
  #################################################

  Ui <- Sigma_gamma %*% t(Zi) %*% Vi_inv
  gamma_i <- Ui %*% resid_i

  #################################################
  # Posterior covariance of random effects:
  #
  # Var(γ_i | Y_i) =
  # Σ_γ − Σ_γ Z_i^T V_i^{-1} Z_i Σ_γ
  #################################################

  Sigma_Ui <- Sigma_gamma - Ui %*% Zi %*% Sigma_gamma

  #################################################
  # Second moment:
  # E[γ_i γ_i^T | Y_i]
  # Needed to update Σ_γ in the M-step
  #################################################

  gamma_cov_i <- gamma_i %*% t(gamma_i) + Sigma_Ui

  list(
    gamma_i = gamma_i,
    gamma_cov_i = gamma_cov_i,
    Sigma_Ui = Sigma_Ui
  )
}

```


# Main Function

This our main function. Important note that the $\hat\Sigma_{\gamma}$ estimates will have the intercept as the last column/row of this covariance matrix.

```{r main_function}

#########################################################
# fastmix_iterative()
# -------------------------------------------------------
# Iterative EM-style mixed model estimation for
# longitudinal data:
#
# - Uses GLS for fixed effects
# - Uses BLUPs for random effects
# - Updates Σ_γ and σ² via moment equations



# IT SHOULD BE NOTED THAT GIVE HOW THIS IS SET UP THE COVARIANCE MATRIX
# OF THE RANDOM EFFECTS HAS THE INTERCEPT AS THE LAST COLUMN!
#########################################################

fastmix_iterative <- function(Y, Des, random = "all", iter = 100) {

  Y = as.matrix(Y)
  N <- length(Y)
  
  # Fixed-effect covariates (excluding ID column)
  covariates <- colnames(Des)[-1]
  p <- length(covariates)
  
  # If all covariates are random effects
  if(length(random) == 1 && random == "all") {
    random <- 1:p
  }
  
  # Random-effects design matrix (ID + random covariates)
  DZ <- Des[, c(1, random + 1)]
  dz <- as.data.frame(DZ)
  colnames(dz)[1] <- "ID"
  
  # Number of subjects
  m <- length(unique(Des[,1]))
  ID <- sort(unique(Des[,1]))
  

  # Build index list mapping subjects to rows

  number <- dz %>% 
    dplyr::group_by(ID) %>% 
    dplyr::summarize(num = dplyr::n(), .groups = "drop") %>% 
    dplyr::mutate(start = dplyr::lag(cumsum(num), default = 0), 
                  end = cumsum(num))
  
  number <- cumsum(number$num)
  number <- c(0, number)
  
  idx_list <- lapply(seq_len(m), function(i) {
    (number[i] + 1):number[i + 1]
  })
  

  # Initial parameter estimates
  #################################################

  # Initial fixed effects via OLS
  coef.fix <- lm.fit(x = as.matrix(Des[, -1]), y = as.matrix(Y))$coefficients
  
  # Initial residual variance
  resid_global <- Y - as.matrix(Des[, -1]) %*% coef.fix
  sigma2_init <- as.numeric(var(resid_global))
  
  # Initial random-effects covariance
  p_rand <- length(random)
  Sigma_gamma_init <- diag(0.1, p_rand)
  
  # Storage across EM iterations
  beta_history <- vector("list", iter)
  Sigma_gamma_history <- vector("list", iter)
  gamma_history <- vector("list", iter)
  sigma2_history <- vector("list", iter)
  
  
  
  
  #################################################
  # EM iterations
  #################################################

  for(it in 1:iter) {
    
    # Subject-specific covariance inverses and GLS pieces
    subject_results <- lapply(seq_len(m), function(i) {
      subject_fun(i, Y, Des, DZ, Sigma_gamma_init, sigma2_init, idx_list)
    })
    
    #################################################
    # M-step: fixed effects via GLS
    #################################################

    p1_sum <- Reduce("+", lapply(subject_results, `[[`, "p1"))
    p2_sum <- Reduce("+", lapply(subject_results, `[[`, "p2"))
    beta_wls <- solve(p1_sum) %*% p2_sum
    
    #################################################
    # E-step: random effects
    #################################################

    gamma_out <- lapply(seq_len(m), function(i) {
      gamma_fun(
        i = i,
        subject_results = subject_results,
        Y = Y,
        Des = Des,
        beta = beta_wls,
        Sigma_gamma = Sigma_gamma_init
      )
    })
    
    #################################################
    # M-step: update Σ_γ
    #################################################

    Sigma_gamma_new <- Reduce("+", lapply(gamma_out, `[[`, "gamma_cov_i")) / m
    
    # Enforce symmetry and positive definiteness
    Sigma_gamma_new <- (Sigma_gamma_new + t(Sigma_gamma_new)) / 2
    eig <- eigen(Sigma_gamma_new)
    eig$values[eig$values < 1e-6] <- 1e-6
    Sigma_gamma_new <- eig$vectors %*% diag(eig$values) %*% t(eig$vectors)
    
    #################################################
    # M-step: update σ²
    #################################################
    
    # Sum the squares first and we take the average after
    ssq <- 0
    for (i in 1:m) {
      Xi <- subject_results[[i]]$Xi
      Yi <- subject_results[[i]]$Yi
      Zi <- subject_results[[i]]$Zi
      g  <- gamma_out[[i]]$gamma_i
      Sigma_Ui <- gamma_out[[i]]$gamma_cov_i - g %*% t(g)
    
      # Expected squared residual
      resid_mean <- Yi - Xi %*% beta_wls - Zi %*% g
      ssq <- ssq +
        sum(resid_mean^2) +
        sum(diag(Zi %*% Sigma_Ui %*% t(Zi)))
    }

    sigma2_new <- ssq / N
    
    #################################################
    # Store iteration results
    #################################################

    beta_history[[it]] <- beta_wls
    Sigma_gamma_history[[it]] <- Sigma_gamma_new
    gamma_history[[it]] <- lapply(gamma_out, `[[`, "gamma_i")
    sigma2_history[[it]] <- sigma2_new
    
    # Update parameters for next iteration
    Sigma_gamma_init <- Sigma_gamma_new
    sigma2_init <- sigma2_new
  }
  
  #################################################
  # Return full estimation history
  #################################################
  
  return(list(
    beta_history = beta_history,
    Sigma_gamma_history = Sigma_gamma_history,
    gamma_history = gamma_history,
    sigma2_history = sigma2_history
  ))
}

#Important note that the $\hat\Sigma_{\gamma}$ estimates will have the intercept
#as the last column/row of this covariance matrix.


```


# Data Generating Process

We create simple DGP with no outliers.


```{r func_dgp}



generate_longitudinal <- function(
  n_subjects = 50,
  n_timepoints = 5,
  n_covariates = 2,
  cov_trends = rep(0, 3),
  beta = rep(1, 3),
  random_intercept_sd = 0.5,
  random_slope_sd = rep(0.5, 3),
  noise_sd = 1,
  include_intercept = TRUE,
  baseline_intercept = 0
) {
  
  # --- checks ---
  if(length(cov_trends) != n_covariates) stop("cov_trends must match n_covariates")
  if(length(beta) != n_covariates) stop("beta must match n_covariates")
  if(length(random_slope_sd) != n_covariates) stop("random_slope_sd must match n_covariates")
  
  # Storage
  rows <- n_subjects * n_timepoints
  df <- data.frame(
    subject = rep(NA, rows),
    time = rep(NA, rows),
    Y = rep(NA, rows),
    matrix(NA, nrow = rows, ncol = n_covariates)
  )
  colnames(df)[4:(3+n_covariates)] <- paste0("X", 1:n_covariates)
  
  # --- Generate random effects ---
  b0 <- rnorm(n_subjects, 0, random_intercept_sd)
  random_slopes <- sapply(random_slope_sd, function(sd) rnorm(n_subjects, 0, sd))
  colnames(random_slopes) <- paste0("b_", paste0("X",1:n_covariates))
  
  # --- Fill dataset ---
  row_counter <- 1
  
  for (subj in 1:n_subjects) {
    for (t in 1:n_timepoints) {
      
      # Generate covariates
      X <- rnorm(n_covariates, mean = cov_trends * (t-1), sd = 1)
      
      # Linear predictor
      intercept <- if(include_intercept) baseline_intercept else 0
      
      rand_eff <- sum(random_slopes[subj, ] * X)
      
      Y <- intercept + 
        b0[subj] + 
        sum(beta * X) + 
        rand_eff + 
        rnorm(1, 0, noise_sd)
      
      # Write row
      df[row_counter, ] <- c(subj, t, Y, X)
      row_counter <- row_counter + 1
    }
  }
  
  # --- True subject-specific effects ---
  # fixed + random slopes
  # true_slopes <- sweep(random_slopes, 2, beta, "+")
  # colnames(true_slopes) <- paste0("slope_X", 1:n_covariates)
  # 
  # true_effects <- data.frame(
  #   subject = 1:n_subjects,
  #   intercept = baseline_intercept + b0,
  #   true_slopes
  # )
  # 
  # --- Return EVERYTHING useful ---
  return(list(
    data = df,
    random_effects = data.frame(
      subject = 1:n_subjects,
      rand_intercept = b0,
      random_slopes)
    # ,
    # true_effects = true_effects,
    # params = list(
    #   beta = beta,
    #   random_intercept_sd = random_intercept_sd,
    #   random_slope_sd = random_slope_sd,
    #   noise_sd = noise_sd,
    #   cov_trends = cov_trends,
    #   baseline_intercept = baseline_intercept
    # )
  ))
}
```

## Running DGP

We are generating 1000 datasets of one variable X1 with a $\beta_1$ = 1 and $\beta_1$ = 0 = 2
Random intercept SD = 1
Random Slope X1 SD = 0.25
$\sigma^2$ = 1

```{r}

n_sims <- 1000
sim_list <- vector("list", n_sims)
simple_re_list <- vector("list", n_sims)
true_effects_list <- vector("list", n_sims)
random_intercept_sd = 1 # random intercept SD
random_slope_sd = c(0.25) #random slope SD


for (i in 1:n_sims) {
  set.seed(123 + i)
  
# Generate one longitudinal dataset according to the DGP
  sim <- generate_longitudinal(
    n_subjects = 50,        # number of subjects
    n_timepoints = 10,     # number of repeated measures per subject
    n_covariates = 1,      # number of time-varying covariates
    cov_trends = c(2.25),  # linear time trend in the covariate
    beta = c(1),           # fixed-effect slope
    random_intercept_sd = random_intercept_sd,     # SD of random intercept
    random_slope_sd = random_slope_sd,   # SD of random slope
    noise_sd = 1,                # residual error SD
    include_intercept = TRUE,    # include fixed intercept
    baseline_intercept = 2       # population-level intercept
  )
  
  # Store the simulated longitudinal data frame
  sim_list[[i]] <- sim$data
  
  # Store the true subject-specific random effects
  # (used to assess recovery of Σ_γ)
  simple_re_list[[i]] <- sim$random_effects
  
  # Store true subject-specific effects (if enabled)
  true_effects_list[[i]] <- sim$true_effects
}

```


#### Sanity check

This is just to make sure my functions are running on a singular dataset before i run on all datasets.
```{r sanity_check}


df_simple = sim_list[[335]]
  RAND_LIST = c(1,2)
  df_simple$X00 = 1
  Des = df_simple[,c(1,4,5)]
  y_values = df_simple[3]
  
  
  random = RAND_LIST
  DZ <- Des[, c(1, random + 1)]
  robust = "FastMix"
  min.cond.num = 1e-6
  Y = as.matrix(y_values)
  bias = 2
  independent = TRUE
  test = 1
  
fm_test = fastmix_iterative(Y = y_values,Des= Des, random = c(1,2),iter = 50)

lmer_model <- lmer(Y ~ 1+ X1 + (1 + X1| subject), data = df_simple)
summary(lmer_model)

```


```{r Wrapper_for_running}
############################################################
# Wrapper to run FastMix on a single longitudinal dataset
############################################################

run_fastmix_on_dataset <- function(df, RAND_LIST = c(1,2), iter = 50) {
  
  # Add an explicit intercept column
  # X00 acts as the fixed intercept in the design matrix
  df$X00 <- 1
  
  # Construct the design matrix (Des)
  # Des is used internally by FastMix to build X_i and Z_i
  Des <- df[, c(1, 4, 5)]
  
  # Extract response variable Y
  y_values <- df[3]
  Y <- as.matrix(y_values)
  
  # Indices of random-effect columns in Des
  random <- RAND_LIST
  
 
  out <- fastmix_iterative(
    Y = Y,
    Des = Des,
    random = random,
    iter = iter
  )

  return(out)
}

```

# Running on all Datasets

## FM
We run our Fastmix on 1000 datasets

```{r parallel_code}
plan(multisession, workers = parallel::detectCores() - 2)
# Start timer
start_time <- Sys.time()
print(start_time)

with_progress({
  p <- progressor(along = sim_list)

  results_list <- future_lapply(sim_list, function(df) {
    p()                              # advance progress bar
    run_fastmix_on_dataset(df)       # run function
  })
})

# End timer
end_time <- Sys.time()

# Time elapsed
elapsed <- end_time - start_time
print(elapsed)

saveRDS(results_list, "RDS_RESULTS_SIMPLE_ONE_COV_FM.RDS")
```


## LMER

Here we run lmer on the 1000 datasets.

```{r lmer_simple_DGP}

run_lmer_on_dataset <- function(df) {
  
  df$Y <- as.numeric(df$Y)
  
  lmer_model <- lmer(
    Y ~ 1 + X1 + (1 + X1| subject),
    data = df,
    REML = TRUE
  )
  
  return(lmer_model)
}




plan(multisession, workers = parallel::detectCores() - 2)

start_time <- Sys.time()
print(start_time)

with_progress({
  p <- progressor(along = sim_list)

  lmer_results_list <- future_lapply(sim_list, function(df) {
    p()
    run_lmer_on_dataset(df)
  })
})
end_time <- Sys.time()
time_needed = end_time-start_time
print(time_needed)


saveRDS(lmer_results_list, "RESULTS_SIMPLE_ONE_COV_LMER.RDS")

```


# Plots

## FASTMIX PLOTS

### Beta Plot
Plot of the Beta MSE over 50 iterations



```{r read_in_saved_results}
results_list = readRDS("RDS_RESULTS_SIMPLE_ONE_COV_FM.RDS")
lmer_results_list = readRDS("RESULTS_SIMPLE_ONE_COV_LMER.RDS")
random_intercept_sd = 1 # random intercept SD
random_slope_sd = c(0.25) #random slope SD
```


```{r plots_simple_new_fast, fig.height=8, fig.width=8}
# ------------------------------
# 1. True beta vector
# ------------------------------

true_beta <- c(1, 2)

# ------------------------------
# 2. Function to compute MSE per iteration
# ------------------------------
compute_error_path <- function(beta_history, true_beta) {
  sapply(beta_history, function(b) {
    b <- as.numeric(b[,1])   # convert 4x1 matrix to vector
    mean((b - true_beta)^2)# MSE
    #sqrt(sum((b - true_beta)^2))# norm
    
    # To use Euclidean norm instead, uncomment the line below:
     #sqrt(sum((b - true_beta)^2))
  })
}

# ------------------------------
# 3. Compute MSE trajectories for all datasets
# ------------------------------
error_list <- lapply(results_list, function(res) {
  compute_error_path(res$beta_history, true_beta)
})

# ------------------------------
# 4. Plotting all trajectories
# ------------------------------
# Start with first dataset to create plot frame
plot(error_list[[1]], type = "l", lwd = 1.5,
     xlab = "Iteration", ylab = "MSE",
     ylim = range(unlist(error_list)),  # ensure all lines fit
     main = "Convergence of β Estimates Across Datasets")

# Overlay remaining datasets with semi-transparent lines
for(i in 2:length(error_list)) {
  lines(error_list[[i]], col = rgb(0,0,0,0.2))
}

# ------------------------------
# 5. Plot average MSE across datasets
# ------------------------------
avg_err <- Reduce("+", error_list) / length(error_list)
lines(avg_err, col = "red", lwd = 3)

# Optional: add legend
legend("topright", legend = c("Individual datasets", "Average"),
       col = c(rgb(0,0,0,0.2), "red"), lwd = c(1.5,3))

```


### Sigma Gamma Plot

```{r sig_gam_plot, fig.height=8, fig.width=8}

# True Σγ from DGP
Sigma_gamma_true <- diag(c(random_slope_sd^2,random_intercept_sd^2))

# Function to compute Frobenius norm errors over iterations
compute_sigma_gamma_error <- function(Sigma_gamma_history, Sigma_true) {
  sapply(Sigma_gamma_history, function(Sigma_est) {
    #sqrt(sum((Sigma_est - Sigma_true)^2))  # Frobenius norm
     mean((Sigma_est - Sigma_true)^2)      # optional: MSE instead
  })
}

# Compute for all datasets
sigma_gamma_error_list <- lapply(results_list, function(res) {
  compute_sigma_gamma_error(res$Sigma_gamma_history, Sigma_gamma_true)
})

# Plot first dataset to create frame
plot(sigma_gamma_error_list[[1]], type = "l", lwd = 1.5,
     xlab = "Iteration", ylab = "MSE Err",
     ylim = c(0,1),
     main = "Convergence of Σγ Estimates Across Datasets")

# Overlay remaining datasets
for(i in 2:length(sigma_gamma_error_list)) {
  lines(sigma_gamma_error_list[[i]], col = rgb(0,0,0,0.2))
}

# Average error across datasets
avg_sigma_err <- Reduce("+", sigma_gamma_error_list) / length(sigma_gamma_error_list)
lines(avg_sigma_err, col = "red", lwd = 3)

# Optional legend
legend("topright", legend = c("Individual datasets", "Average"),
       col = c(rgb(0,0,0,0.2), "red"), lwd = c(1.5,3))



```

### sigma^2 plots

Trajectories of the $\sigma^2$.

```{r traj_sigma2, fig.height=8, fig.width=8}

# Convert sigma2_history list of each dataset into numeric vector
sigma2_trajectories <- lapply(results_list, function(res) {
  unlist(res$sigma2_history)
})
# Plot the first dataset trajectory
plot(sigma2_trajectories[[1]], type = "l", lwd = 1.5,
     xlab = "Iteration", ylab = expression(sigma^2),
     ylim = range(unlist(sigma2_trajectories)),
     main = expression("Trajectory of"~sigma^2~"across all datasets"))

# Overlay remaining datasets
for(i in 2:length(sigma2_trajectories)) {
  lines(sigma2_trajectories[[i]], col = rgb(0,0,0,0.2))
}

# Average trajectory across datasets
avg_sigma2 <- Reduce("+", sigma2_trajectories) / length(sigma2_trajectories)
lines(avg_sigma2, col = "red", lwd = 3)

# True σ² from DGP
true_sigma2 <- 1
abline(h = true_sigma2, col = "blue", lty = 2)

# Legend
legend("topright", legend = c("Individual datasets", "Average", "True σ²"),
       col = c(rgb(0,0,0,0.2), "red", "blue"), lwd = c(1.5,3,2), lty = c(1,1,2))


```





### gamma indiv

Average of the mean MSE across 50 subjects per dataset. This plot takes a while to run

```{r gamma_individual, fig.height=8, fig.width=8}
# ------------------------------
# 1. Function to compute average error per iteration for dataset
# ------------------------------
compute_avg_subject_gamma_error <- function(gamma_history, true_re_df) {
  # gamma_history: list of iterations, each element is list of subject vectors
  # true_re_df: data frame with columns: rand_intercept, b_X1, b_X2, b_X3
  n_iter <- length(gamma_history)
  n_subjects <- nrow(true_re_df)
  
  sapply(1:n_iter, function(it) {
    # For each subject, compute squared error
    errors <- sapply(1:n_subjects, function(subj_idx) {
      gamma_est <- as.numeric(gamma_history[[it]][[subj_idx]])
      gamma_true <- as.numeric(true_re_df[subj_idx, -1])  # drop subject column
      mean((gamma_est - gamma_true)^2)  # MSE per subject
      #sqrt(sum((gamma_est - gamma_true)^2))  #norm
    })
    mean(errors)  # average over all subjects
  })
}

# ------------------------------
# 2. Compute average errors for all datasets
# ------------------------------

gamma_error_list <- lapply(1:length(results_list), function(i) {
  compute_avg_subject_gamma_error(
    gamma_history = results_list[[i]]$gamma_history,
    true_re_df = simple_re_list[[i]]
  )
})

# ------------------------------
# 3. Plotting all datasets together
# ------------------------------
# Plot first dataset to create plot frame
plot(gamma_error_list[[1]], type = "l", lwd = 1.5,
     xlab = "Iteration", ylab = "Average MSE per subject",
     ylim = range(unlist(gamma_error_list)),
     main = "Convergence of γ Estimates Across Datasets")

# Overlay remaining datasets with semi-transparent lines
for(i in 2:length(gamma_error_list)) {
  lines(gamma_error_list[[i]], col = rgb(0,0,0,0.2))
}

# Average across datasets
avg_gamma_err <- Reduce("+", gamma_error_list) / length(gamma_error_list)
lines(avg_gamma_err, col = "red", lwd = 3)

# Add legend
legend("topright", legend = c("Individual datasets", "Average"),
       col = c(rgb(0,0,0,0.2), "red"), lwd = c(1.5, 3))

```



## Final Results



```{r fin_results}



true_beta = c(2,1) # intercept beta comes first
true_sigma_gamma = diag(c(1,0.25)^2)
## FASTMIX

N_results<- length(results_list)

# Extract final betas from each dataset
final_betas <- lapply(1:N_results, function(i) t(results_list[[i]]$beta_history[[50]]))
beta_mat <- do.call(rbind, final_betas)
beta_mean <- colMeans(beta_mat)




# Extract final Sigma_gamma from each dataset
final_Sigmas <- lapply(1:N_results, function(i) results_list[[i]]$Sigma_gamma_history[[50]])
Sigma_array <- simplify2array(final_Sigmas)
Sigma_gamma_mean <- apply(Sigma_array, c(1, 2), mean)


final_sigma2 <- sapply(1:N_results, function(i) results_list[[i]]$sigma2_history[[50]])
sigma2_mean <- mean(final_sigma2)


# Current covariance matrix
S <- Sigma_gamma_mean

# Desired order: move row/col 4 to the front
new_order <- c(2, 1)

# Reordered covariance matrix
S_reordered <- S[new_order, new_order]
Sigma_gamma_mean = S_reordered
len_beta <- length(beta_mean)
beta_mean <- c(beta_mean[len_beta], beta_mean[-len_beta])

cat("\n\n\ FOR FASTMIX \n")
cat("\n\n The Sigma^2 estimate: \n", sigma2_mean)
cat("\n\n The Sigma Gamma Cov Matrix estimate: \n\n\n")
print(Sigma_gamma_mean)
cat( "\n\n with MSE: ", mean((Sigma_gamma_mean-true_sigma_gamma)^2 ))
cat("\n\n The beta estimate: \n")
cat(beta_mean, "\n\n with MSE: ", mean((beta_mean-true_beta)^2 ))


### LMER

N_results <- length(lmer_results_list)
fixed_effects_list <- lapply(1:N_results, function(i) fixef(lmer_results_list[[i]]))
Sigma_gamma_list <- lapply(1:N_results, function(i) as.matrix(VarCorr(lmer_results_list[[i]])$subject))
sigma2_list <- sapply(1:N_results, function(i) sigma(lmer_results_list[[i]])^2)


fixed_mat <- do.call(rbind, lapply(fixed_effects_list, t))
beta_mean_lmer <- colMeans(fixed_mat)

Sigma_array <- simplify2array(Sigma_gamma_list)
Sigma_gamma_mean_lmer <- apply(Sigma_array, c(1, 2), mean)
sigma2_mean_lmer <- mean(sigma2_list)


cat("\n\\n\n\n\n\ FOR LMER \n")
cat("\n\n The Sigma^2 estimate: \n", sigma2_mean_lmer)
cat("\n\n The Sigma Gamma Cov Matrix estimate: \n\n\n")
print(Sigma_gamma_mean_lmer)
cat("\n\n with MSE: ", mean((Sigma_gamma_mean_lmer-true_sigma_gamma))^2)
cat("\n\n The beta estimate: \n")
cat(beta_mean_lmer, "\n\n with MSE: ", mean((beta_mean_lmer-true_beta)^2),"\n\n")





# # Current covariance matrix
# S <- Sigma_gamma_mean
# 
# # Desired order: move row/col 4 to the front
# new_order <- c(4, 1, 2, 3)
# 
# # Reordered covariance matrix
# S_reordered <- S[new_order, new_order]
# Sigma_gamma_mean = S_reordered





mse_all <- sapply(seq_along(results_list), function(i) {
  
  # true REs reordered to match fastmix ordering
  true_gamma <- simple_re_list[[i]][ , c("b_X1","rand_intercept")]
  
  # estimated REs at iteration 50
  est_gamma_list <- results_list[[i]]$gamma_history[[50]]
  est_gamma <- do.call(rbind, est_gamma_list)
  
  # MSE per subject
  mse_subject <- rowMeans((est_gamma - true_gamma)^2)
  
  # average MSE for this dataset
  mean(mse_subject)
})

# Final average across datasets
cat("\n\n The mean MSE across all datasets for gamma estimates for FM: ")
print(mean(mse_all))

mse_lmer_all <- sapply(seq_along(lmer_results_list), function(i) {
  
  # TRUE random effects reordered to match lmer ordering (X1, X2, X3, intercept)
  true_gamma <- simple_re_list[[i]][ , 
                                      c("b_X1","rand_intercept")]
  
  # EST random effects from lmer
  est_gamma <- as.matrix(ranef(lmer_results_list[[i]])[[1]])
  est_gamma <- est_gamma[ , c("X1","(Intercept)")]
  # Subject-level MSE
  mse_subject <- rowMeans((est_gamma - true_gamma)^2)
  
  # Dataset-level MSE
  mean(mse_subject)
})

# Overall average across datasets
overall_mse_lmer <- mean(mse_lmer_all)

cat("\n\n The mean MSE across all datasets for gamma estimates for LMER: ")
print(overall_mse_lmer)
cat("\n\n\n\n TRUE SIGMA_GAMMA: \n")

true_sigma_gamma
cat("\n FastMix SIGMA_GAMMA: \n")
Sigma_gamma_mean
cat("\n LMER SIGMA_GAMMA: \n")
Sigma_gamma_mean_lmer

```
# OUTLIER DGP
We create a dataset with 3 variables.
$\beta_0$ =2
$\beta_1$ = 1
$\beta_2$ = 2
$\beta_3$ = -1

Random Intercept = 1
Random Slope SD = 0.25, 0.75, 0.5 [$X_1$,$X_2$,$X_3$]

Here we create a dataset with outliers of 30% in variable $X_1$ of the "mix" variant.
This shifts the random slope, random intercept, Y-values and X-values by a magnitude of its SD.

```{r dgp_outlier_v2}
generate_longitudinal_with_outliers <- function(
  # --- data-generating parameters (identical to generate_longitudinal) ---
  n_subjects = 50,
  n_timepoints = 10,
  n_covariates = 3,
  cov_trends = c(2.25, -1.25, 0.65),
  beta = c(1, 2, -1),
  random_intercept_sd = 1,
  random_slope_sd = c(0.25, 0.75, 0.5),
  noise_sd = 1,
  include_intercept = TRUE,
  baseline_intercept = 2,
  
  # --- outlier parameters (your additions) ---
  outlier_prop = 0,
  outlier_type = "mix",    # slope, intercept, covariate, response, mix
  outlier_magnitude = 2,
  outlier_covariates = NULL,
  seed = NULL
) {
  if (!is.null(seed)) set.seed(seed)
  
  # ---- SAME checks as original ----
  if(length(cov_trends) != n_covariates) stop("cov_trends must match n_covariates")
  if(length(beta) != n_covariates) stop("beta must match n_covariates")
  if(length(random_slope_sd) != n_covariates) stop("random_slope_sd must match n_covariates")
  
  rows <- n_subjects * n_timepoints
  df <- data.frame(
    subject = rep(NA, rows),
    time = rep(NA, rows),
    Y = rep(NA, rows),
    matrix(NA, nrow = rows, ncol = n_covariates)
  )
  colnames(df)[4:(3+n_covariates)] <- covariate_names <- paste0("X", 1:n_covariates)
  
  # ---- random effects (same as original) ----
  b0 <- rnorm(n_subjects, 0, random_intercept_sd)
  random_slopes <- sapply(random_slope_sd, function(sd) rnorm(n_subjects, 0, sd))
  colnames(random_slopes) <- paste0("b_", covariate_names)
  
  # ---- main data loop (identical to original) ----
  row_counter <- 1
  for (subj in 1:n_subjects) {
    for (t in 1:n_timepoints) {
      
      X <- rnorm(n_covariates, mean = cov_trends * (t-1), sd = 1)
      
      intercept_val <- if(include_intercept) baseline_intercept else 0
      
      re <- sum(random_slopes[subj, ] * X)
      Y <- intercept_val + b0[subj] + sum(beta * X) + re + rnorm(1, 0, noise_sd)
      
      df[row_counter, ] <- c(subj, t, Y, X)
      row_counter <- row_counter + 1
    }
  }
  
  # ---- Outlier logic ----
  outlier_subjects <- sample(n_subjects, size = round(outlier_prop * n_subjects))
  df$outlier_flag <- df$subject %in% outlier_subjects
  
  if (length(outlier_subjects) > 0) {
    
    if (is.null(outlier_covariates)) outlier_covariates <- covariate_names
    
    for (sid in outlier_subjects) {
      rows_sid <- df$subject == sid
      
      # extreme slope effects → modify random slopes
      if (outlier_type %in% c("slope","mix")) {
        for (j in seq_len(n_covariates)) {
          jump <- outlier_magnitude * random_slope_sd[j]
          random_slopes[sid, j] <- random_slopes[sid, j] + sample(c(-1,1),1)*jump
        }
      }
      
      # extreme intercepts
      if (outlier_type %in% c("intercept","mix")) {
        b0[sid] <- b0[sid] + sample(c(-1,1),1) * outlier_magnitude * random_intercept_sd
      }
      
      # covariate contamination
      if (outlier_type %in% c("covariate","mix")) {
        for (v in outlier_covariates) {
          sdv <- sd(df[[v]])
          df[rows_sid, v] <- df[rows_sid, v] +
            sample(c(-1,1),1) * outlier_magnitude * sdv
        }
      }
      
      # response contamination
      if (outlier_type %in% c("response","mix")) {
        df$Y[rows_sid] <- df$Y[rows_sid] +
          rnorm(sum(rows_sid), 0, outlier_magnitude * noise_sd)
      }
    }
  }
  
  # ---- Random effects output (same format as original) ----
  re_df <- data.frame(
    subject = 1:n_subjects,
    rand_intercept = b0,
    random_slopes
  )
  
  return(list(
    data = df,
    random_effects = re_df
  ))
}


# edirty_sim <- generate_longitudinal_with_outliers(
#   outlier_prop = 0.30,
#   outlier_type = "mix",
#   outlier_magnitude = 3,
#   outlier_covariates = "X1",
# 
#   # everything below matches your clean DGP exactly:
#   n_subjects = 50,
#   n_timepoints = 10,
#   n_covariates = 3,
#   cov_trends = c(2.25, -1.25, 0.65),
#   beta = c(1, 2, -1),
#   random_intercept_sd = 1,
#   random_slope_sd = c(0.25, 0.75, 0.5),
#   noise_sd = 1,
#   include_intercept = TRUE,
#   baseline_intercept = 2
# )
# 
# df = edirty_sim$data


n_sims = 1000
sim_list_outlier = list()
simple_re_list_outlier = list()
for (i in 1:n_sims) {
  set.seed(123 + i)
  
  sim <- generate_longitudinal_with_outliers(
    outlier_prop = 0.30,
  outlier_type = "mix",
  outlier_magnitude = 2,
  outlier_covariates = "X1",

  # everything below matches your clean DGP exactly:
  n_subjects = 50,
  n_timepoints = 10,
  n_covariates = 3,
  cov_trends = c(2.25, -1.25, 0.65),
  beta = c(1, 2, -1),
  random_intercept_sd = 1,
  random_slope_sd = c(0.25, 0.75, 0.5),
  noise_sd = 1,
  include_intercept = TRUE,
  baseline_intercept = 2
  )
  
  sim_list_outlier[[i]] <- sim$data
  simple_re_list_outlier[[i]] <- sim$random_effects
}





```



### sanity check p2

```{r sanity_check_p2}
  df = sim_list_outlier[[1]]
  df$X00 <- 1
  df <- df[, c(names(df)[1:(ncol(df)-2)], "X00", names(df)[ncol(df)-1])]
  Des <- df[, c(1, 4, 5, 6, 7)]
  y_values <- df[3]
  
  Y <- as.matrix(y_values)
  random <- RAND_LIST
  
  out <- fastmix_iterative(
    Y = Y,
    Des = Des,
    random = random,
    iter = 30
  )

lmer_model <- lmer(Y ~ 1+ X1+ X2 + X3 + (1 + X1+X2+X3| subject), data = df)
summary(lmer_model)

```

# Running on Outliers


## Fastmix

```{r running on outliers}
run_fastmix_on_dataset_outlier <- function(df, RAND_LIST = c(1,2,3,4), iter = 50) {
  
  df$X00 <- 1
  df <- df[, c(names(df)[1:(ncol(df)-2)], "X00", names(df)[ncol(df)-1])]
  Des <- df[, c(1, 4, 5, 6, 7)]
  y_values <- df[3]
  
  Y <- as.matrix(y_values)
  random <- RAND_LIST
  
  out <- fastmix_iterative(
    Y = Y,
    Des = Des,
    random = random,
    iter = iter
  )

  return(out)
}

library(progressr)

plan(multisession, workers = parallel::detectCores() - 1)
# Start timer
start_time <- Sys.time()
print(start_time)

with_progress({
  p <- progressor(along = sim_list_outlier)

  results_list <- future_lapply(sim_list_outlier, function(df) {
    p()                              # advance progress bar
    run_fastmix_on_dataset_outlier(df)       # run your function
  })
})

end_time <- Sys.time()

# Time elapsed
elapsed <- end_time - start_time
print(elapsed)

saveRDS(results_list, "RESULTS_FM_OUTLIER.RDS")
```


## LMER


```{r lmer_running_all_outlier_df}

run_lmer_on_dataset <- function(df) {
  

  df$Y <- as.numeric(df$Y)
  
  lmer_model <- lmer(
    Y ~ 1 + X1 +X2 + X3 + (1 + X1 + X2 + X3| subject),
    data = df,
    REML = TRUE
  )
  
  return(lmer_model)
}



plan(multisession, workers = parallel::detectCores() - 1)
start_time <- Sys.time()
print(start_time)

with_progress({
  p <- progressor(along = sim_list_outlier)

  lmer_results_list <- future_lapply(sim_list_outlier, function(df) {
    p()
    run_lmer_on_dataset(df)
  })
})
end_time <- Sys.time()
time_needed = end_time - start_time
print(time_needed)


saveRDS(lmer_results_list, "RESULTS_LMER_OUTLIER.RDS")


```

# Plots Outliers


```{r read_in_saved_results_outlier }
results_list = readRDS("RESULTS_FM_OUTLIER.RDS")
lmer_results_list = readRDS("RESULTS_LMER_OUTLIER.RDS")

```

## Beta Plot

```{r plots_simple_new_fast, fig.height=8, fig.width=8}

# ------------------------------
# 1. True beta vector
# ------------------------------
true_beta <- c(1, 2, -1, 2)

# ------------------------------
# 2. Function to compute MSE per iteration
# ------------------------------
compute_error_path <- function(beta_history, true_beta) {
  sapply(beta_history, function(b) {
    b <- as.numeric(b[,1])   # convert 4x1 matrix to vector
    mean((b - true_beta)^2)# MSE
    # To use Euclidean norm instead, uncomment the line below:
    # sqrt(sum((b - true_beta)^2))
  })
}

# ------------------------------
# 3. Compute MSE trajectories for all datasets
# ------------------------------
error_list <- lapply(results_list, function(res) {
  compute_error_path(res$beta_history, true_beta)
})

# ------------------------------
# 4. Plotting all trajectories
# ------------------------------
# Start with first dataset to create plot frame
plot(error_list[[1]], type = "l", lwd = 1.5,
     xlab = "Iteration", ylab = "MSE ERR",
     ylim = range(unlist(error_list)),  # ensure all lines fit
     main = "Convergence of β Estimates Across Datasets")

# Overlay remaining datasets with semi-transparent lines
for(i in 2:length(error_list)) {
  lines(error_list[[i]], col = rgb(0,0,0,0.2))
}

# ------------------------------
# 5. Plot average MSE across datasets
# ------------------------------
avg_err <- Reduce("+", error_list) / length(error_list)
lines(avg_err, col = "red", lwd = 3)

# Optional: add legend
legend("topright", legend = c("Individual datasets", "Average"),
       col = c(rgb(0,0,0,0.2), "red"), lwd = c(1.5,3))


```

## Sigma Gamma Plot

```{r sigma_gamma_outlier, fig.height=8, fig.width=8}
# True Σγ from DGP
random_intercept_sd <- 1
random_slope_sd <- c(0.25, 0.75, 0.5)
Sigma_gamma_true <- diag(c(random_slope_sd^2,random_intercept_sd^2))

# Function to compute Frobenius norm errors over iterations
compute_sigma_gamma_error <- function(Sigma_gamma_history, Sigma_true) {
  sapply(Sigma_gamma_history, function(Sigma_est) {
    #sqrt(sum((Sigma_est - Sigma_true)^2))  # Frobenius norm
     mean((Sigma_est - Sigma_true)^2)      # optional: MSE instead
  })
}

# Compute for all datasets
sigma_gamma_error_list <- lapply(results_list, function(res) {
  compute_sigma_gamma_error(res$Sigma_gamma_history, Sigma_gamma_true)
})

# Plot first dataset to create frame
plot(sigma_gamma_error_list[[1]], type = "l", lwd = 1.5,
     xlab = "Iteration", ylab = "MSE Err",
     ylim = range(c(0,0.3)),
     main = "Convergence of Σγ Estimates Across Datasets")

# Overlay remaining datasets
for(i in 2:length(sigma_gamma_error_list)) {
  lines(sigma_gamma_error_list[[i]], col = rgb(0,0,0,0.2))
}

# Average error across datasets
avg_sigma_err <- Reduce("+", sigma_gamma_error_list) / length(sigma_gamma_error_list)
lines(avg_sigma_err, col = "red", lwd = 3)

# Optional legend
legend("topright", legend = c("Individual datasets", "Average"),
       col = c(rgb(0,0,0,0.2), "red"), lwd = c(1.5,3))

```


## Sigma Gamma TOL
```{r tolerance_sig_gamma_outlier, fig.height=8, fig.width=8}
### Sigma Gamma Convergence (Difference between Iterations)

# Function to compute Error between consecutive iterations
compute_sigma_gamma_tolerance <- function(Sigma_gamma_history) {
  n_iter <- length(Sigma_gamma_history)
  
  # Loop from 2 to end (since we need t and t-1)
  sapply(2:n_iter, function(i) {
    diff_matrix <- Sigma_gamma_history[[i]] - Sigma_gamma_history[[i-1]]
    #sqrt(sum(diff_matrix^2)) # Frobenius norm of the difference
    mean(diff_matrix^2)
  })
}

# 1. Compute tolerance path for all datasets
sigma_gamma_tol_list <- lapply(results_list, function(res) {
  compute_sigma_gamma_tolerance(res$Sigma_gamma_history)
})

# 2. Plotting
# Note: The x-axis will be 2:50 (49 points) because we lose the first point
iterations_x <- 2:length(results_list[[1]]$Sigma_gamma_history)

# Setup the empty plot frame
plot(iterations_x, sigma_gamma_tol_list[[1]], type = "n", # 'n' for no points yet
     ylim = range(0.01,0),
     xlab = "Iteration", ylab = " MSE (t vs t-1)",
     main = "Convergence Tolerance of Σγ (Consecutive Diff)")

# Overlay all datasets
for(i in 1:length(sigma_gamma_tol_list)) {
  lines(iterations_x, sigma_gamma_tol_list[[i]], col = rgb(0,0,0,0.2))
}

# 3. Calculate and plot the average trajectory
avg_sigma_tol <- Reduce("+", sigma_gamma_tol_list) / length(sigma_gamma_tol_list)
lines(iterations_x, avg_sigma_tol, col = "blue", lwd = 3)

legend("topright", legend = c("Individual Traces", "Average Tolerance"),
       col = c(rgb(0,0,0,0.2), "blue"), lwd = c(1,3))
```





## sigma^2 plots


```{r traj_sigma2_outlier}

# Convert sigma2_history list of each dataset into numeric vector
sigma2_trajectories <- lapply(results_list, function(res) {
  unlist(res$sigma2_history)
})
# Plot the first dataset trajectory
plot(sigma2_trajectories[[1]], type = "l", lwd = 1.5,
     xlab = "Iteration", ylab = expression(sigma^2),
     ylim = range(10,0),
     main = expression("Trajectory of"~sigma^2~"across all datasets"))

# Overlay remaining datasets
for(i in 2:length(sigma2_trajectories)) {
  lines(sigma2_trajectories[[i]], col = rgb(0,0,0,0.2))
}

# Average trajectory across datasets
avg_sigma2 <- Reduce("+", sigma2_trajectories) / length(sigma2_trajectories)
lines(avg_sigma2, col = "red", lwd = 3)

# True σ² from DGP
true_sigma2 <- 1
abline(h = true_sigma2, col = "blue", lty = 2)

# Legend
legend("topright", legend = c("Individual datasets", "Average", "True σ²"),
       col = c(rgb(0,0,0,0.2), "red", "blue"), lwd = c(1.5,3,2), lty = c(1,1,2))


```


## Final Results



```{r fin_results}
true_beta = c(2,1,2,-1)
random_intercept_sd <- 1
random_slope_sd <- c(0.25, 0.75, 0.5)
true_sigma_gamma <- diag(c(random_intercept_sd^2,random_slope_sd^2))
## FASTMIX

N_results<- length(results_list)

# Extract final betas from each dataset
final_betas <- lapply(1:N_results, function(i) t(results_list[[i]]$beta_history[[50]]))
beta_mat <- do.call(rbind, final_betas)
beta_mean <- colMeans(beta_mat)




# Extract final Sigma_gamma from each dataset
final_Sigmas <- lapply(1:N_results, function(i) results_list[[i]]$Sigma_gamma_history[[50]])
Sigma_array <- simplify2array(final_Sigmas)
Sigma_gamma_mean <- apply(Sigma_array, c(1, 2), mean)


final_sigma2 <- sapply(1:N_results, function(i) results_list[[i]]$sigma2_history[[50]])
sigma2_mean <- mean(final_sigma2)


# Current covariance matrix
S <- Sigma_gamma_mean

# Desired order: move row/col 4 to the front
new_order <- c(4,1,2,3)

# Reordered covariance matrix
S_reordered <- S[new_order, new_order]
Sigma_gamma_mean = S_reordered
len_beta <- length(beta_mean)
beta_mean <- c(beta_mean[len_beta], beta_mean[-len_beta])

cat("\n\n\ FOR FASTMIX \n")
cat("\n\n The Sigma^2 estimate: \n", sigma2_mean)
cat("\n\n The Sigma Gamma Cov Matrix estimate: \n\n\n")
print(Sigma_gamma_mean)
cat( "\n\n with MSE: ", mean((Sigma_gamma_mean-true_sigma_gamma)^2 ))
cat("\n\n The beta estimate: \n")
cat(beta_mean, "\n\n with MSE: ", mean((beta_mean-true_beta)^2 ))


### LMER

N_results <- length(lmer_results_list)
fixed_effects_list <- lapply(1:N_results, function(i) fixef(lmer_results_list[[i]]))
Sigma_gamma_list <- lapply(1:N_results, function(i) as.matrix(VarCorr(lmer_results_list[[i]])$subject))
sigma2_list <- sapply(1:N_results, function(i) sigma(lmer_results_list[[i]])^2)


fixed_mat <- do.call(rbind, lapply(fixed_effects_list, t))
beta_mean_lmer <- colMeans(fixed_mat)

Sigma_array <- simplify2array(Sigma_gamma_list)
Sigma_gamma_mean_lmer <- apply(Sigma_array, c(1, 2), mean)
sigma2_mean_lmer <- mean(sigma2_list)


cat("\n\n\n\n\n\n\ FOR LMER \n")
cat("\n\n The Sigma^2 estimate: \n", sigma2_mean_lmer)
cat("\n\n The Sigma Gamma Cov Matrix estimate: \n\n\n")
print(Sigma_gamma_mean_lmer)
cat("\n\n with MSE: ", mean((Sigma_gamma_mean_lmer-true_sigma_gamma))^2)
cat("\n\n The beta estimate: \n")
cat(beta_mean_lmer, "\n\n with MSE: ", mean((beta_mean_lmer-true_beta)^2),"\n\n")





# # Current covariance matrix
# S <- Sigma_gamma_mean
# 
# # Desired order: move row/col 4 to the front
# new_order <- c(4, 1, 2, 3)
# 
# # Reordered covariance matrix
# S_reordered <- S[new_order, new_order]
# Sigma_gamma_mean = S_reordered





mse_all <- sapply(seq_along(results_list), function(i) {
  
  # true REs reordered to match fastmix ordering
  true_gamma <- simple_re_list_outlier[[i]][ , c("b_X1","b_X2","b_X3","rand_intercept")]
  
  # estimated REs at iteration 50
  est_gamma_list <- results_list[[i]]$gamma_history[[50]]
  est_gamma <- do.call(rbind, est_gamma_list)
  
  # MSE per subject
  mse_subject <- rowMeans((est_gamma - true_gamma)^2)
  
  # average MSE for this dataset
  mean(mse_subject)
})

# Final average across datasets
cat("\n\n The mean MSE across all datasets for gamma estimates for FM: ")
print(mean(mse_all))

mse_lmer_all <- sapply(seq_along(lmer_results_list), function(i) {
  
  # TRUE random effects reordered to match lmer ordering (X1, X2, X3, intercept)
  true_gamma <- simple_re_list_outlier[[i]][ , c("b_X1","b_X2","b_X3","rand_intercept")]
  
  # EST random effects from lmer
  est_gamma <- as.matrix(ranef(lmer_results_list[[i]])[[1]])
  est_gamma <- est_gamma[ , c("X1","X2","X3","(Intercept)")]
  # Subject-level MSE
  mse_subject <- rowMeans((est_gamma - true_gamma)^2)
  
  # Dataset-level MSE
  mean(mse_subject)
})

# Overall average across datasets
overall_mse_lmer <- mean(mse_lmer_all)

cat("\n\n The mean MSE across all datasets for gamma estimates for LMER: ")
print(overall_mse_lmer)


cat("\n\n\n\n TRUE SIGMA_GAMMA: \n")

true_sigma_gamma
cat("\n FastMix SIGMA_GAMMA: \n")
Sigma_gamma_mean
cat("\n LMER SIGMA_GAMMA: \n")
Sigma_gamma_mean_lmer

```